Unshuffling Permutations

### 2015-09-16 article.tex
matching

\begin{proof}[of Lemma~\ref{lemma:matching}]
    \qed
\end{proof}

@@

\begin{proof}[of Lemma~\ref{lemma:at most one edge monotone}]
  \begin{figure}[t!]
    \centering
    \subfigure[%
      An increasing pattern before and above a decreasing pattern.
    ]{%
      \begin{tikzpicture}
        [
          scale=0.325,
          label/.style={anchor=base}
        ]
        % increasing pattern
        \draw[step=1cm,black!30,ultra thin,fill=black!10] (-0.2,5.8) grid (5.2,11.2);
        \foreach \x/\y in {0/6,1/7,2/8,3/9,4/10,5/11} {
            \draw [fill=black] (\x,\y) circle (0.1);
        }
        % decreasing pattern
        \draw[step=1cm,black!30,ultra thin,fill=black!10] (5.8,-0.2) grid (11.2,5.2);
        \foreach \x/\y in {6/5,7/4,8/3,9/2,10/1,11/0} {
            \draw [fill=black] (\x,\y) circle (0.1);
        }
        % labels
        \node [label] (a) at (1,5) {$a$};
        \node [label] (b) at (4,5) {$b$};
        \node [label] (ap) at (7,-1) {$a'$};
        \node [label] (bp) at (10,-1) {$b'$};
        \node (nu1) at (2.5,4) {$\sigma_1$};
        \node (nu2) at (8.5,-2) {$\sigma_2$};
        % crossings edges
        \draw [line width=1pt]
        (1,7) .. controls +(0,12) and +(0,10) .. (7,4);
        \draw [line width=1pt]
        (4,10) .. controls +(0,12) and +(0,10) .. (10,1);
      \end{tikzpicture}
      \label{subfig:increasing before and aboce decreasing}
    }% end subfigure
    \qquad\qquad\qquad
    \subfigure[%
      A decreasing pattern before and below an increasing pattern.
    ]{%
    \begin{tikzpicture}
      [
        scale=0.325,
        label/.style={anchor=base}
      ]
      % decreasing pattern
      \draw[step=1cm,black!30,ultra thin,fill=black!10] (-0.2,-0.2) grid (5.2,5.2);
      \foreach \x/\y in {0/5,1/4,2/3,3/2,4/1,5/0} {
          \draw [fill=black] (\x,\y) circle (0.1);
      }
      % decreasing pattern
      \draw[step=1cm,black!30,ultra thin,fill=black!10] (5.8,5.8) grid (11.2,11.2);
      \foreach \x/\y in {6/6,7/7,8/8,9/9,10/10,11/11} {
          \draw [fill=black] (\x,\y) circle (0.1);
      }
      % labels
      \node [label] (a) at (1,-1) {$a$};
      \node [label] (b) at (4,-1) {$b$};
      \node [label] (ap) at (7,5) {$a'$};
      \node [label] (bp) at (10,5) {$b'$};
      \node (nu1) at (2.5,-2) {$\sigma_1$};
      \node (nu2) at (8.5,4) {$\sigma_2$};
      % crossings edges
      \draw [line width=1pt]
      (1,4) .. controls +(0,10) and +(0,12) .. (7,7);
      \draw [line width=1pt]
      (4,1) .. controls +(0,10) and +(0,12) .. (10,10);
      \end{tikzpicture}
      \label{subfig:decreasing before and below increasing}
    }% end subfigure
    \caption{\label{fig:subfig:no (nu_1, nu_2)-edge}%
      Lemma~\ref{lemma:at most one edge monotone}:
      Assuming a $(\nu_1,\nu_2)$-edge $(a, b)$ in $\mathcal{M}$.
    }
  \end{figure}
  We first prove \emph{1}.
  Suppose, aiming at a contradiction, that there exist two
  $(\sigma_1, \sigma_2)$-edges $(a, a')$ and $(b, b')$ in $\mathcal{M}$.
  Since $\mathcal{M}$ is containment-free, we may assume $a < b$ and $a' < b'$.
  Therefore, we have
  $\pi(a) < \pi(b)$ (since $\sigma_1$ is increasing) and
  $\pi(a') > \pi(b')$ (since $\sigma_2$ is decreasing),
  which contradicts our hypothesis about $\mathcal{M}$.

@@

There are a number of further directions of investigation in this
general subject. We mention several -~not necessarily new~- open
problems that are, in our opinion, the most interesting. How many
permutations of $S_{2n}$ are squares? How many $(213,231)$-avoiding
permutations of $S_{2n}$ are squares? (Equivalently, how many binary
strings of length $2n$ are squares; see Problem~4
in \cite{Henshall:Rampersad:Shallit:2011})? How hard is the problem of
deciding whether a $(213,231)$-avoiding permutation is a square
(see Problem~4 in \cite{Henshall:Rampersad:Shallit:2011})?) Given two
permutations $\pi$ and $\sigma$, how hard is the problem of deciding
whether $\pi$ is the union of two patterns order-isomorphic to $\sigma$?
\smallskip

>>>>

\begin{proof}[of Lemma~\ref{lemma:matching}]
  The proof is, naturally enough, a slight extension of
  Lemma~1 in \cite{Rizzi:Vialette:CSR:2013} and
  is added here in full for the sake of completeness.

  (\emph{1} $\Rightarrow$ \emph{2}).
  Let $\sigma$ be a permutation such that $\pi$ is the union of two
  patterns that are order-isomorphic to $\sigma$.
  Let $2n = |\pi|$ (and hence $|\sigma| = n$).
  Fix the occurrences in $\pi$ of the two disjoint patterns that
  are order-isomorphic to $\sigma$ and
  write
  $I^1 = \{i^1_1, i^1_2, \ldots, i^1_n\}$,
  $i^1_1 < i^1_2 < \ldots < i^1_n$, for the positions in $\pi$
  of the first pattern order-ismorphic to $\sigma$ and
  $I^2 = \{i^2_1, i^2_2, \ldots, i^2_n\}$,
  $i^2_1 < i^2_2 < \ldots < i^2_n$, for the positions in $\pi$
  of the second pattern order-ismorphic to $\sigma$.
  Let $G_\pi$ the linear clique associated to $\pi$ with
  vertices $\VS{G_\pi} =
  \{i^1_j : 1 \leq j \leq n\} \cup \{i^2_j : 1 \leq j \leq n\}$.
  It is easily seen that
  $\mathcal{M} = \{\{i^1_j, i^2_j\} : 1 \leq j \leq n\}$
  is a subset of $\ES{G_\pi}$.
  Furthermore, $\mathcal{M}$ is a perfect matching since
  $I^1 \cap I^2 = \emptyset$,
  and $I^1 \cup I^2 = \{1, 2,  \ldots 2n\}$.
  It is also containment-free.
  Indeed, if it were not the case then there would exist two edges
  $e = \{i^1_j, i^2_j\}$ and
  $e' = \{i^1_k, i^2_k\}$, $j < k$,
  in $\mathcal{M}$ such that
  $i^1_j < i^1_k < i^2_k < i^2_j$ or
  $i^1_j < i^2_k < i^1_k < i^2_j$.
  This is a contradiction since
  $i^2_k > i^2_j$ if $k > j$.
  Finally, for any two distinct edges
  $e = \{\i^1_j, i^2_j\}$ and
  $e' = \{i^1_k, i^2_k\}$ of $\mathcal{M}$,
  we have $\pi(i^1_j) < \pi(i^2_j)$ if and only if
  $\pi(i^1_k) < \pi(i^2_k)$ since we are comparing
  in both cases two elements (at position $j$ and $k$)
  in two patterns that are order-isomorphic to $\sigma$.

  (\emph{2} $\Rightarrow$ \emph{1}).
  Let $I^1 = \{i^1_j : \exists \{i_j, i_k\} \in \mathcal{M}
  \;\text{with}\; j < k\}$, and
  $I^2 = \{i^2_k : \exists \{i_j, i_k\} \in \mathcal{M}
  \;\text{with}\; j < k\}$.
  Informally, $I^1$ (resp. $I^2$) is the set of the left (resp. right) endpoints
  of the edges in $\mathcal{M}$ assuming the positions $1, 2, \ldots, 2$
  are drawn along an horizontal line.
  Without loss of generality,
  assume $i^1_1 < i^1_2 < \ldots < i^1_n$.
  % and
  % $I^2 = \{i^2_1, i^2_2, \ldots, i^2_n\}$.
  Let
  $p^1 = \pi(i^1_1) \pi(i^1_2) \ldots \pi(i^1_n)$ and
  $p^2 = \pi(i^2_1) \pi(i^2_2) \ldots \pi(i^2_n)$ be the two patterns
  from $\pi$ induced by $I^1$ and $I^2$, respectivelly.
  Clearly $p^1$ and $p^2$ are disjoint in $\pi$ (since $\mathcal{M}$ is
  a matching) and cover $\pi$ (since $\mathcal{M}$ is perfect).
  We also claim that $p^1$ and $p^2$ are order-isomorphic.
  Suppose, aiming at a contradiction, that this is not the case.
  Therefore, there exist two positions $j$ and $k$, say $j < k$, such that
  $\pi(i^1_j) < \pi(i^2_j)$ and $\pi(i^1_k) > \pi(i^2_k)$, or
  $\pi(i^1_j) < \pi(i^2_j)$ and $\pi(i^1_k) > \pi(i^2_k)$.
  But since $\mathcal{M}$ is containment-free, it follows that
  both $\{i^1_j, i^2_j\}$ and $\{i^1_k, i^2_k\}$.
  This is a contradiction since
  for any two disctinct edges
  $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  we have $\pi(a) < \pi(a')$ if and only if $\pi(b) < \pi(b')$.
  \qed
\end{proof}

@@

\begin{proof}[of Lemma~\ref{lemma:at most one edge monotone}]
  \begin{figure}[t!]
    \centering
    \subfigure[%
      An increasing pattern before and above a decreasing pattern.
    ]{%
      \begin{tikzpicture}
        [
          scale=0.325,
          label/.style={anchor=base}
        ]
        % increasing pattern
        \draw[step=1cm,black!30,ultra thin,fill=black!10] (-0.2,5.8) grid (5.2,11.2);
        \foreach \x/\y in {0/6,1/7,2/8,3/9,4/10,5/11} {
            \draw [fill=black] (\x,\y) circle (0.1);
        }
        % decreasing pattern
        \draw[step=1cm,black!30,ultra thin,fill=black!10] (5.8,-0.2) grid (11.2,5.2);
        \foreach \x/\y in {6/5,7/4,8/3,9/2,10/1,11/0} {
            \draw [fill=black] (\x,\y) circle (0.1);
        }
        % labels
        \node [label] (a) at (1,5) {$a$};
        \node [label] (b) at (4,5) {$b$};
        \node [label] (ap) at (7,-1) {$a'$};
        \node [label] (bp) at (10,-1) {$b'$};
        \node (nu1) at (2.5,4) {$\sigma_1$};
        \node (nu2) at (8.5,-2) {$\sigma_2$};
        % crossings edges
        \draw [line width=1pt]
        (1,7) .. controls +(0,12) and +(0,10) .. (7,4);
        \draw [line width=1pt]
        (4,10) .. controls +(0,12) and +(0,10) .. (10,1);
      \end{tikzpicture}
      \label{subfig:increasing before and aboce decreasing}
    }% end subfigure
    \qquad\qquad\qquad
    \subfigure[%
      A decreasing pattern before and below an increasing pattern.
    ]{%
    \begin{tikzpicture}
      [
        scale=0.325,
        label/.style={anchor=base}
      ]
      % decreasing pattern
      \draw[step=1cm,black!30,ultra thin,fill=black!10] (-0.2,-0.2) grid (5.2,5.2);
      \foreach \x/\y in {0/5,1/4,2/3,3/2,4/1,5/0} {
          \draw [fill=black] (\x,\y) circle (0.1);
      }
      % decreasing pattern
      \draw[step=1cm,black!30,ultra thin,fill=black!10] (5.8,5.8) grid (11.2,11.2);
      \foreach \x/\y in {6/6,7/7,8/8,9/9,10/10,11/11} {
          \draw [fill=black] (\x,\y) circle (0.1);
      }
      % labels
      \node [label] (a) at (1,-1) {$a$};
      \node [label] (b) at (4,-1) {$b$};
      \node [label] (ap) at (7,5) {$a'$};
      \node [label] (bp) at (10,5) {$b'$};
      \node (nu1) at (2.5,-2) {$\sigma_1$};
      \node (nu2) at (8.5,4) {$\sigma_2$};
      % crossings edges
      \draw [line width=1pt]
      (1,4) .. controls +(0,10) and +(0,12) .. (7,7);
      \draw [line width=1pt]
      (4,1) .. controls +(0,10) and +(0,12) .. (10,10);
      \end{tikzpicture}
      \label{subfig:decreasing before and below increasing}
    }% end subfigure
    \caption{\label{fig:subfig:no (nu_1, nu_2)-edge}%
      Illustration of Lemma~\ref{lemma:at most one edge monotone}.
    }
  \end{figure}
  We first prove \emph{1}.
  Suppose, aiming at a contradiction, that there exist two
  $(\sigma_1, \sigma_2)$-edges $(a, a')$ and $(b, b')$ in $\mathcal{M}$.
  Since $\mathcal{M}$ is containment-free, we may assume $a < b$ and $a' < b'$.
  Therefore, we have
  $\pi(a) < \pi(b)$ (since $\sigma_1$ is increasing) and
  $\pi(a') > \pi(b')$ (since $\sigma_2$ is decreasing),
  which contradicts our hypothesis about $\mathcal{M}$.

@@

There are a number of further directions of investigation in this
general subject. We mention several -~not necessarily new~- open
problems that are, in our opinion, the most interesting. How many
permutations of $S_{2n}$ are squares? How many $(213,231)$-avoiding
permutations of $S_{2n}$ are squares? (Equivalently, how many binary
strings of length $2n$ are squares; Problem~4
in \cite{Henshall:Rampersad:Shallit:2011})? How hard is the problem of
deciding whether a $(213,231)$-avoiding permutation is a square
(Problem~4 in \cite{Henshall:Rampersad:Shallit:2011},
see also \cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013})?)
% Given two
% permutations $\pi$ and $\sigma$, how hard is the problem of deciding
% whether $\pi$ is the union of two patterns order-isomorphic to $\sigma$?
\smallskip
